Deep learning approaches on musical assisted-orchestration: an evaluation study

### 2020-05-07 paper/latex/paper.tex
minor edits

When training a model, the number of samples $n$ to be used in each combination is specified. To create the training data, $n$ instruments are randomly selected from the list of specified instruments, and then a TinySOL sample is randomly selected for each chosen instrument. This gives $n$ different samples, each of a different intrument. These samples are combined to form a soundfile that has each sample playing simultaneously. The melspectrogram is taken of this combination, creating the features to be fed to the model. We created the melspectrogram by \textbf{(Alejandro can you write this of how you generated the melspectrograms using the STFT?)}

@@

Our idea is to perform orchestration using deep neural networks trained for classification. By generating input data that is combinations of different instruments and pitches, the network learns how to take a sound that is a complex combination of notes and timbres and deconstruct it into its original parts. When a target sound is then fed to the network, it attempts to classify which samples are present. By taking the samples that have the highest probability of being in the target, an orchestrated solution can be created.

@@

The first deep model we tested was a Convolutional Neural Network (CNN). CNN shows good performance on classification assignment for its strong ability to extract spacial features, so we implemented a basic CNN with three convolutional layers and two fully connected layers. Each convolutional layer is followed by a BatchNorm layer, a ReLU activation layer and a 2 * 2  MaxPool layer with a stride of 2. The kernel size is 3 * 3 with a stride of 1 and a padding of 1. The number of filters are 8, 16, 32. This setting is simple but we can make a good comparison between it and the baseline.

@@

Given that audio is one kind of sequential data, it is easy to consider Long Short-Term Memory network(LSTM) as test model. In our setting, we added one LSTM layer with 32 outputs units following three convolutional layers. The convolutional layers did not change the sequence relationship. Instead, it scaled down the size of feature maps. After LSTM layer, we reshaped the output into the same size before inputing. Then we implemented another convolutional layer with the same structure and kernel size. Finally, we flattened out the outputs and fed them into fully connected layers with Dropout layer. The probability of an element to be zeroed is 0.5 and we took sigmoid as activation function. We used binary classification for each class and treated them independently.

@@

We evaluated our orchestrations both qualitatively and quantitatively. Qualitative inspection was done through an acoustic inspection of the solution, paying close attention to timbre and pitch. For targets that had harmonic content, it was noted if the partials present in the target were also represented in the orchestrated solution. For example, one of the samples of a bell had partials that represented a C sharp minor chord. For a target that contained specific notes, the solution from ResNet contained both the note in the target and its partials: a sample that was two octaves higher, the fifth an octave up, and a minor third two octaves up (mix\_ObA4\_BnC3).

@@

We find that the CNN and ResNet give similar accuracies during training, but perform differently when tasked with orchestrating targets. Overall, CNN seems to better emulate the timbre in its orchestrations, where ResNet is better for recreating the pitch(es) of the target.

>>>>

When training a model, the number of samples $n$ to be used in each combination is specified. To create the training data, $n$ instruments are randomly selected from the list of specified instruments, and then a TinySOL sample is randomly selected for each chosen instrument. This gives $n$ different samples, each of a different intrument. These samples are combined to form a soundfile that has each sample playing simultaneously. The melspectrogram is taken of this combination, creating the features to be fed to the model. We created the melspectrogram by \textbf{(Alejandro can you write this of how you generated the melspectrograms using the STFT?)}. We chose to use the melspectrogram because of its ability to represent timbre, a key part of assisted orchestration.

@@

Our idea is to perform orchestration using deep neural networks trained for classification. By generating input data that is a mixture of different instruments and pitches, the network learns how to take a sound that is a complex combination of notes and timbres and deconstruct it into its original parts. When a target sound is then fed to the network, it attempts to classify which samples would best represent the target. By taking the samples that have the highest probability of being in the target, an orchestrated solution can be created.

@@

The first deep model we tested was a Convolutional Neural Network (CNN). CNN shows good performance on classification assignment for its strong ability to extract spatial features, so we implemented a basic CNN with three convolutional layers and two fully connected layers. Each convolutional layer is followed by a BatchNorm layer, a ReLU activation layer and a 2 * 2  MaxPool layer with a stride of 2. The kernel size is 3 * 3 with a stride of 1 and a padding of 1. The number of filters are 8, 16, 32. This setting is simple but we can make a good comparison between it and the baseline.

@@

Given that audio is one kind of sequential data, it is easy to consider Long Short-Term Memory network (LSTM) as test model. In our setting, we added one LSTM layer with 32 outputs units following three convolutional layers. The convolutional layers did not change the sequence relationship. Instead, it scaled down the size of feature maps. After LSTM layer, we reshaped the output into the same size before inputing. Then we implemented another convolutional layer with the same structure and kernel size. Finally, we flattened out the outputs and fed them into fully connected layers with Dropout layer. The probability of an element to be zeroed is 0.5 and we took sigmoid as activation function. We used binary classification for each class and treated them independently.

@@

We evaluated our orchestrations both qualitatively and quantitatively. Qualitative inspection was done through an acoustic inspection of the solution, paying close attention to timbre and pitch. For targets that had harmonic content, it was noted if the partials present in the target were also represented in the orchestrated solution.

@@

We find that the CNN and ResNet give similar accuracies during training, but perform differently when tasked with orchestrating targets. Overall, CNN seems to better emulate the timbre in its orchestrations, where ResNet is better for recreating the pitch of the target. For a target that contained specific notes, the solution from ResNet contained both the note in the target and its partials: a sample that was two octaves higher, the fifth an octave up, and a minor third two octaves up (mix\_ObA4\_BnC3). Clearly, ResNet was able to identify the partials of the target and recreate them in the solution.
